---
title: "msleep"
author: "Morgan Purcell"
date: "11/22/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


```{r}
library(tidyverse)
library(ggplot2)

```

```{r}
dim(msleep)
names(msleep)
str(msleep)
sample_n(msleep, 10)

```



1. What's the average mammal sleep time and how do mammals with the highest and lowest sleep times compare to this number?

```{r}
summarise(msleep,mean(sleep_total))

arrange(msleep, sleep_total)

arrange(msleep, desc(sleep_total))

```

The average time spent sleeping of all animals in a day is 10.4 hours. It is interesting to see that giraffes only sleep 1.9 hours a day while little brown bats sleep 19.9 hours a day. Upon viewing the data, it appears heavier mammals tend to sleep less than lighter mammals with some exceptions. 



2. Do lighter mammals tend to sleep more while heavier mammals tend to sleep less?

```{r}
ggplot(data = msleep, aes(x = sleep_total, y = bodywt)) +
  geom_point() + geom_smooth() +
  labs(title = "Body Weights Effect on Mammals Total Sleep Time",
       x = "Total Sleep Time (Hours)",
       y = "Body Weight (KG)")

```
The graph indicates that there is some evidence to suggest that heavier mammals sleep less than lighter mammals. Mammals below the 5 hour sleep mark seem to be some of the heaviest mammals overall. The graph does hit a plateau after the 8 hour sleep time however. This could mean that there are is just an abundance of lighter mammals above the 8 hour sleep time. There are two outlines above the 2000 kilogram body weight. It would be interesting to identify these mammals and test the hypothesis again, excluding these two mammals which would also yield a closer look at the spread of the data.



3. What are the two outlines over 2000 kilograms.

```{r}
filter(msleep, bodywt > 2000)

```
The mammals over 2000 kilograms are the Asian and African elephants. It is interesting to see that the African elephant weighs much more than the Asian elephant and sleeps 0.6 hours less.



4. How would a graph to compare the relationship between body weight and sleep times look without the African and Asian Elephants.

```{r}
below2000 <- filter(msleep, bodywt < 2000)

ggplot(data = below2000, aes(x = sleep_total, y = bodywt)) +
  geom_point() + geom_smooth() +
  labs(title = "Body Weights Effect on Mammals Total Sleep Time",
       subtitle = "Excluding mammals over 2000 KG",
       x = "Total Sleep Time (Hours)",
       y = "Body Weight (KG)")

```
The data can be seen more clearly with the exclusion of the African and Asian elephants. There is significant dip in the graph below 7 hours sleep total meaning some very heavy mammals do sleep very little but the graph does plateau after this meaning both heavy and lighter mammals sleep times are similar. Could it be that the majority of mammals in the data are considered light?



5. What is the average body weight of the mammals and how many mammals are above and below this average?

```{r}
summarise(msleep, mean(bodywt))

heavym <- filter(msleep, bodywt > 166.1363)
count(heavym)

lightm <- filter(msleep, bodywt < 166.1363)
count(lightm)

```
The average body weight of all mammals is 166.1363. There are 74 mammals below the average while there are only 9 mammals above this average. The data set doesn't seem to contain enough "heavy" mammals to conclude that heavier mammals tend to sleep less than lighter mammals. Would brain weight have a similar effect on total sleep time compared to body wight?



6. How does a mammals brain weight effect total sleep time.

```{r}
sample_n(msleep, 10)

ggplot(data = msleep, aes(x = sleep_total, y = bodywt)) +
  geom_point() + geom_smooth() +
  labs(title = "Body Weights Effect on Total Sleep Time",
       x = "Total Sleep Time (Hours)",
       y = "Body Weight (KG)")

ggplot(data = msleep, aes(x = sleep_total, y = brainwt)) +
  geom_point() + geom_smooth() +
  labs(title = "Brain Weights Effect on Mammals Total Sleep Time",
       x = "Total Sleep Time (Hours)",
       y = "Brain Weight (KG)")

```
Brain weight does seem to have the same effect on total sleep time. There could be other variables which influence mammal sleep times such as conservation and diet.



7. How does a mammals diet (carnivore, herbivore, insectivore, and omnivore) effect sleep time.

```{r}
ggplot(data = msleep, aes(x = vore, y = sleep_total)) +
  geom_boxplot() +
  labs(title = "Diets Effect on Mammals Total Sleep Time",
       x = "Diet",
       y = "Total Sleep Time (Hours)")

```
Carnivores, herbivores, and omnivore mammals all have very similar sleep times with some outlines for omnivores. Insectivores differ from the others though as they have a much higher average sleep time. 



8. How does a mammals conservation effect sleep time.

```{r}
ggplot(data = msleep, aes(x = conservation, y = sleep_total)) +
  geom_boxplot() +
  labs(title = "Conservations Effect on Mammals Total Sleep Time",
       x = "Conservation",
       y = "Total Sleep Time (Hours)")

```
The data is very dispersed and shows different conversations yield varied sleep times. It would be interesting to group together all the non domesticated mammals and compare their average total sleep time to that of the domesticated mammals.



9. Compare the average total sleep times for domesticated vs non-domesticated mammals

```{r}
dom <- filter(msleep, conservation == "domesticated")
summarise(dom, mean(sleep_total))

non_dom <- filter(msleep, conservation != "domesticated")
summarise(non_dom, mean(sleep_total))

```
The averages show that domesticated mammals sleep less than non-domesticated mammals. The average sleep time for domesticated animals is a very interesting figure at 7.58 hours. I see two explanations for this figure. The first being that domesticated animals, which live alongside humans, have grown accustomed to the human sleep times of 7 to 9 hours a day. A second explanation could be that many non-domesticated animals have to spend time and energy hunting their prey which means they need more rest.


To conclude there are some variables that seem to influence a mammals total sleep time. There isn't a sufficient amount of data or more specifically, a sufficient amount of heavy mammals to conclude that body weight and brain weight influence total sleep times. It was interesting to see that only insectivores differed in total sleep times when comparing diets. Each conservation varied in total sleep time but when comparing domesticated and non-domesticated mammals, domesticated animals sleep around 3 hours less. 